---
title: "Getting Started with krtexas"
author: "Sithija Manage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with krtexas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
***

# 📦 Introduction

The **krtexas** package implements **Kernel Regression with Tree-EXploring AggregationS (KR-TEXAS)**, a method for **nonparametric regression with structured predictors**.

Many datasets naturally exhibit **hierarchical structure**, such as:

* Microbiome taxonomies (species → genus → family)
* Brain regions (voxels → regions → systems)
* Financial sectors and industries
* Geographic hierarchies

KR-TEXAS leverages this structure to **learn both**:

* which features matter
* at what level of aggregation they matter

***

# ⚙️ Model Overview

KR-TEXAS solves the penalized optimization problem:

$$
\min_{\gamma} \; L_n(\gamma) + \lambda \sum_v w_v \gamma_v
$$

where:

* $L\_n(\gamma)$ is a leave-one-out kernel regression loss
* $\gamma\_v$ controls whether node $v$ is selected
* $w\_v$ are adaptive weights constructed from data

The method proceeds in three stages:

1. **Stage 1**: Fit model with $\lambda = 0$
2. **Stage 2**: Construct adaptive weights
3. **Stage 3**: Fit penalized model with cross-validation

***

# 🚀 Basic Usage

```{r}
library(krtexas)
```

## Simulated Example

We simulate a simple dataset with a tree structure.

```{r}
set.seed(1)

n <- 200
p <- 6

# Features
X <- matrix(rnorm(n * p), n, p)

# True relationship (depends only on sum of first 2 features)
Y <- 3*(X[,1] + X[,2])^2 + rnorm(n, sd = 0.01)

# Example tree matrix A (T x p)
A <- rbind(  
  diag(p),          # leaves
  c(1,1,0,0,0,0),  # group 1
  c(0,0,1,1,0,0),  # group 2
  c(1,1,1,1,0,0)  # group 3
)
```

Because the true relationship only depends on the *sum* of the *first two* features,
we would like our regression method to set the coefficients of all
other features to 0. Because the sum of the first two features is represented
by group 1 (row number 7), we would like the learned gamma vector to take the 
form c(0,0,0,0,0,0,[nonzero],0,0).

## Fit Model

```{r, results = 'hide', warning = FALSE}
fit <- krtexas_fit(
  X = X,
  Y = Y,
  A = A
)
```

## Inspect Output

```{r}
str(fit)
```

Key components:

* `gammas_learned`: selected features / aggregations
* `lambda.best`: optimal regularization parameter
* `training_loss`: final objective value

Note that as we hoped, all gamma values except for the one corresponding to 
group 1 are set to 0! This means that our estimator properly identified that
the sum of the first two variables are the only ones relevant for predicting
the outcome variable, and the value itself, 43, reflects how predictive changes
in this variable are. 

***

# 🔮 Prediction on New Data

```{r}
X_test <- matrix(rnorm(50 * p), 50, p)

pred <- krtexas_predict(fit, newx = X_test)

head(pred)
```

***

# 📊 Interactive 3D visualization of the learned surface

You can click and drag the plot to inspect it :)

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
library(plotly)

# Grid over X1 and X2
grid_size <- 50

x1_seq <- seq(min(X[,1]), max(X[,1]), length.out = grid_size)
x2_seq <- seq(min(X[,2]), max(X[,2]), length.out = grid_size)

grid_2d <- expand.grid(x1_seq, x2_seq)

# Fix remaining variables at their mean
X_fixed <- matrix(colMeans(X), nrow = nrow(grid_2d), ncol = ncol(X), byrow = TRUE)

X_fixed[,1] <- grid_2d[,1]
X_fixed[,2] <- grid_2d[,2]

# Predictions using existing fit
z_pred <- krtexas_predict(fit, newx = X_fixed)

# Convert to matrix
Z <- matrix(z_pred, nrow = grid_size, byrow = FALSE)

# Plotly surface
plot_ly() %>%
  add_surface(
    x = x1_seq,
    y = x2_seq,
    z = Z,
    colorscale = "Viridis",
    opacity = 0.8
  ) %>%
  add_markers(
    x = X[,1],
    y = X[,2],
    z = Y,
    marker = list(size = 2, color = "black")
  ) %>%
  layout(
    title = "KR-TEXAS Surface with Training Data",
    scene = list(
      xaxis = list(title = "X1"),
      yaxis = list(title = "X2"),
      zaxis = list(title = "Response")
    )
  )
```


# 📖 References

> Manage, S., Wang, Y. S., & Wells, M. T. (2026).  
> *Nonparametric Regression via Tree-Guided Feature Aggregation*.  
> arXiv preprint.

***

# ✅ Summary

KR-TEXAS provides a flexible framework for:

* nonlinear regression
* structured feature selection
* interpretable aggregation

For a full list of parameters, please see the README (https://github.com/sithijamanage/krtexas). 
Feel free to email me with any questions. I should be pretty quick to respond at
ssm255@cornell.edu 🙂!

***
